---
title: "gRate: from raw plate export to clean growth curves"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gRate: from raw plate export to clean growth curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

gRate is the preprocessing layer between your plate reader and growth
model fitting. It reads raw 96-well growth curve exports, maps the plate
layout, flags problematic wells, corrects spatial artifacts, and hands tidy
data to packages like growthcurver or gcplyr. It deliberately does **not** fit
growth models — those packages already do that well.

The whole pipeline is five verbs on one object:

```{r, eval = FALSE}
gr_read("run1.csv") |>
  gr_layout("layout.csv") |>
  gr_qc() |>
  gr_spatial() |>
  gr_export(as = "growthcurver")
```

This vignette walks through each step using the example data bundled with the
package: a synthetic 24-hour OD600 run with realistic problems baked in — two
dead wells, three wells with single-read spikes, and a multiplicative edge
effect.

```{r setup}
library(gRate)
```

## 1. Read

`gr_read()` accepts the two generic layouts plate reader software exports:
**wide** (a time column plus one column per well) and **long** (well / time /
value columns). The layout is auto-detected; well names like `A01` are
normalised to `A1`, and `HH:MM:SS` times are converted to hours.

```{r read}
raw <- system.file("extdata", "growth_wide.csv", package = "gRate")
plate <- gr_read(raw)
plate
```

Everything lives in a single `gr_plate` object: `$data` (tidy tibble),
`$qc` (per-well flags, once QC has run), and `$meta`.

```{r data}
head(plate$data)
```

## 2. Layout

`gr_layout()` joins your experimental design onto the wells. The recommended
format is a long table with a `well` column plus any metadata columns; an
8 × 12 grid mirroring the physical plate also works for one variable at a
time.

```{r layout}
layout <- system.file("extdata", "layout_long.csv", package = "gRate")
plate <- gr_layout(plate, layout)
head(plate$data)
```

### Biological and technical replicates

Two metadata columns are treated specially: `bio_rep` (independent cultures)
and `tech_rep` (the same culture in several wells). Columns already named
that way — as in the example layout — are picked up automatically; otherwise
point the `bio_rep` / `tech_rep` arguments at your own column names:

```{r layout-reps, eval = FALSE}
gr_layout(plate, "layout.csv", bio_rep = "biol_replicate", tech_rep = "tech")
```

Once designated, the replicate structure shows up in the print summary and
technical replicates can be averaged at export time (below).

```{r print-reps}
plate
```

## 3. Quality control

`gr_qc()` flags wells that would corrupt downstream fits: `no_growth`,
`spike`, `drift`, `late_jump`, and `noisy`. Every threshold is an argument
with a sensible default. Flags never delete data — you decide later what to
drop.

```{r qc}
plate <- gr_qc(plate)
plate
subset(plate$qc, flagged)
```

The two plotting functions make the QC results easy to eyeball. The plate
heatmap shows any statistic or flag in its physical position:

```{r plot-plate}
gr_plot_plate(plate, "max_od")
gr_plot_plate(plate, "flagged")
```

Notice the dimmer outer ring in the `max_od` map — that is the edge effect we
will deal with next. The curve view highlights flagged wells in orange:

```{r plot-curves, fig.height = 6}
gr_plot_curves(plate)
```

Zoom in on the flagged wells to see why they were caught:

```{r plot-flagged}
gr_plot_curves(plate, wells = subset(plate$qc, flagged)$well)
```

## 4. Spatial correction (experimental)

Edge wells often read lower — evaporation and temperature gradients are the
usual suspects. `gr_spatial()` estimates row and column effects on a per-well
summary statistic (max OD by default) using Tukey's median polish, then
divides each curve by its estimated bias factor. Flagged wells are excluded
from the estimation but still receive the correction.

```{r spatial}
plate <- gr_spatial(plate)
round(plate$spatial$row_effects, 3)
round(plate$spatial$col_effects, 3)
```

Rows A and H and columns 1 and 12 sit well below 1 — the edge effect — while
the interior is essentially unbiased. After correction the plate map is flat:

```{r spatial-after}
gr_plot_plate(plate, "max_od")
```

Treat this step as a diagnostic aid: median polish captures row/column trends
(corner wells are corrected twice over), and no correction rescues a design
that confounds treatment with plate position. Randomise your layouts. The
uncorrected values remain in `plate$data$value_raw`, and `correct = FALSE`
estimates the effects without touching the data.

## 5. Export

`gr_export()` returns plain data frames in the shape downstream tools expect.
`drop_flagged = TRUE` is the one place gRate removes data — explicitly,
at your request.

```{r export}
tidy <- gr_export(plate)
head(tidy)

gc_input <- as_growthcurver(plate, drop_flagged = TRUE)
gc_input[1:3, 1:5]
```

The wide `gc_input` goes straight into growthcurver:

```{r growthcurver, eval = requireNamespace("growthcurver", quietly = TRUE)}
fits <- growthcurver::SummarizeGrowthByPlate(gc_input)
head(fits)
```

For gcplyr, `as_gcplyr()` produces the long `Well` / `Time` / `Measurements`
shape (metadata columns included).

```{r gcplyr}
head(as_gcplyr(plate, drop_flagged = TRUE))
```

### Averaging technical replicates

With `tech_rep` designated, `collapse_tech = TRUE` averages the wells of each
technical replicate group at every timepoint. Combined with
`drop_flagged = TRUE`, only clean wells enter the average; otherwise `flagged`
tells you when any contributing well was suspect.

```{r collapse}
pooled <- gr_export(plate, collapse_tech = TRUE, drop_flagged = TRUE)
head(pooled)
```

## One-call Quarto report

`gr_report(plate)` renders the bundled Quarto (`.qmd`) template into a
standalone, self-contained HTML report — plate maps, flagged wells with
reasons, thresholds used, curves, and spatial effects — for keeping alongside
the raw export. It needs the Quarto CLI (bundled with recent RStudio).

```{r report, eval = FALSE}
gr_report(plate, file = "run1_qc.html")
```
