---
title: "gRate: from raw plate export to growth rates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gRate: from raw plate export to growth rates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

gRate takes you from a raw 96-well plate reader export to growth parameters
you can trust. It reads the export, maps the plate layout, flags problematic
wells, corrects spatial artifacts, fits growth models (parametric logistic or
nonparametric "easylinear"), and summarises parameters across replicates.
Prefer growthcurver or gcplyr for the fitting? `gr_export()` hands them
clean, QC'd data instead.

The whole pipeline is six verbs on one object:

```{r, eval = FALSE}
gr_read("run1.csv") |>
  gr_layout("layout.csv") |>
  gr_qc() |>
  gr_spatial() |>
  gr_fit() |>
  gr_fit_summary()
```

This vignette walks through each step using the example data bundled with the
package: a synthetic 24-hour OD600 run with realistic problems baked in — two
dead wells, three wells with single-read spikes, and a multiplicative edge
effect.

```{r setup}
library(gRate)
```

## 1. Read

`gr_read()` accepts the two generic layouts plate reader software exports:
**wide** (a time column plus one column per well) and **long** (well / time /
value columns). The layout is auto-detected; well names like `A01` are
normalised to `A1`, and `HH:MM:SS` times are converted to hours.

```{r read}
raw <- system.file("extdata", "growth_wide.csv", package = "gRate")
plate <- gr_read(raw)
plate
```

Everything lives in a single `gr_plate` object: `$data` (tidy tibble),
`$qc` (per-well flags, once QC has run), and `$meta`.

```{r data}
head(plate$data)
```

## 2. Layout

`gr_layout()` joins your experimental design onto the wells. The recommended
format is a long table with a `well` column plus any metadata columns; an
8 × 12 grid mirroring the physical plate also works for one variable at a
time.

```{r layout}
layout <- system.file("extdata", "layout_long.csv", package = "gRate")
plate <- gr_layout(plate, layout)
head(plate$data)
```

### Biological and technical replicates

Two metadata columns are treated specially: `bio_rep` (independent cultures)
and `tech_rep` (the same culture in several wells). Columns already named
that way — as in the example layout — are picked up automatically; otherwise
point the `bio_rep` / `tech_rep` arguments at your own column names:

```{r layout-reps, eval = FALSE}
gr_layout(plate, "layout.csv", bio_rep = "biol_replicate", tech_rep = "tech")
```

Once designated, the replicate structure shows up in the print summary and
technical replicates can be averaged at export time (below).

```{r print-reps}
plate
```

## 3. Quality control

`gr_qc()` flags wells that would corrupt downstream fits: `no_growth`,
`spike`, `drift`, `late_jump`, and `noisy`. Every threshold is an argument
with a sensible default. Flags never delete data — you decide later what to
drop.

```{r qc}
plate <- gr_qc(plate)
plate
subset(plate$qc, flagged)
```

The two plotting functions make the QC results easy to eyeball. The plate
heatmap shows any statistic or flag in its physical position:

```{r plot-plate}
gr_plot_plate(plate, "max_od")
gr_plot_plate(plate, "flagged")
```

Notice the dimmer outer ring in the `max_od` map — that is the edge effect we
will deal with next. The curve view highlights flagged wells in orange:

```{r plot-curves, fig.height = 6}
gr_plot_curves(plate)
```

Zoom in on the flagged wells to see why they were caught:

```{r plot-flagged}
gr_plot_curves(plate, wells = subset(plate$qc, flagged)$well)
```

## 4. Spatial correction (experimental)

Edge wells often read lower — evaporation and temperature gradients are the
usual suspects. `gr_spatial()` estimates row and column effects on a per-well
summary statistic (max OD by default) using Tukey's median polish, then
divides each curve by its estimated bias factor. Flagged wells are excluded
from the estimation but still receive the correction.

```{r spatial}
plate <- gr_spatial(plate)
round(plate$spatial$row_effects, 3)
round(plate$spatial$col_effects, 3)
```

Rows A and H and columns 1 and 12 sit well below 1 — the edge effect — while
the interior is essentially unbiased. After correction the plate map is flat:

```{r spatial-after}
gr_plot_plate(plate, "max_od")
```

Treat this step as a diagnostic aid: median polish captures row/column trends
(corner wells are corrected twice over), and no correction rescues a design
that confounds treatment with plate position. Randomise your layouts. The
uncorrected values remain in `plate$data$value_raw`, and `correct = FALSE`
estimates the effects without touching the data.

## 5. Fit growth models

`gr_fit()` estimates growth parameters per well, QC-aware and on the
spatially corrected values:

* `method = "logistic"` (default) — parametric `nls` fit of the logistic
  model: growth rate `r`, carrying capacity `K`, initial density `N0`,
  tangent lag time, doubling time.
* `method = "easylinear"` — nonparametric, after Hall et al. (2014): rolling
  regressions on log OD; the steepest window passing an R² filter gives the
  maximum per-capita growth rate. Use it when curves are not logistic.

Wells that cannot be fitted (the dead wells here) get `fit_ok = FALSE` and a
`note` — never an error, and never silently dropped.

Want uncertainty on the estimates? `boot = 200` resamples the residuals and
refits per well, adding percentile confidence intervals (`r_lo`/`r_hi`, and
`K_lo`/`K_hi` for the logistic method) to the fit table and to
`gr_results()`. It takes a minute for a full plate, so it is off by default:

```{r fit-boot, eval = FALSE}
plate <- gr_fit(plate, boot = 200)
gr_results(plate, params = c("r", "K"))  # now includes r_lo, r_hi, K_lo, K_hi
```

```{r fit}
plate <- gr_fit(plate)
plate
head(plate$fit)
```

Overlay the fits on the data to judge them by eye, and map any parameter back
onto the plate:

```{r fit-plots, fig.height = 6}
gr_plot_fit(plate, wells = c("A1", "B3", "C5", "E4"))
gr_plot_plate(plate, "r")
```

### The results table

`gr_results()` is the table to carry into your analysis: one row per well,
with your metadata, the parameters you care about, and the QC verdict side by
side. Pick parameters with `params`, drop flagged wells if you want only
clean fits, or write it straight to CSV with `file`.

```{r results}
gr_results(plate)
gr_results(plate, params = c("r", "K"), drop_flagged = TRUE)
```

```{r results-csv, eval = FALSE}
gr_results(plate, file = "run1_growth_parameters.csv")
```

### Summarise across replicates

`gr_fit_summary()` is the step most fitting tools leave to you: it averages
parameters over technical and biological replicates *after* excluding flagged
wells and failed fits. By default it groups by every metadata column except
the replicate identifiers; pass `by` to group differently.

```{r fit-summary}
gr_fit_summary(plate)
gr_fit_summary(plate, by = c("strain", "medium", "bio_rep"))
```

### Compare groups

`gr_compare()` tests whether a parameter differs between strains or
conditions — at the right unit of replication. Technical replicates are
averaged into their biological replicate *before* any test, so wells never
inflate the sample size. Two groups get a Welch t-test, more get Welch's
ANOVA plus Holm-adjusted pairwise comparisons (`method = "kruskal"` for the
nonparametric versions).

```{r compare}
cmp <- gr_compare(plate, what = "r", by = "strain")
cmp
gr_plot_compare(cmp)
```

(The bundled example data were generated with the same growth rate for every
strain, so an honest test finds nothing — exactly as it should.)

## 6. Export

Prefer growthcurver or gcplyr as the fitting engine? `gr_export()` returns
plain data frames in the shape those tools expect.
`drop_flagged = TRUE` is the one place gRate removes data — explicitly,
at your request.

```{r export}
tidy <- gr_export(plate)
head(tidy)

gc_input <- as_growthcurver(plate, drop_flagged = TRUE)
gc_input[1:3, 1:5]
```

The wide `gc_input` goes straight into growthcurver:

```{r growthcurver, eval = requireNamespace("growthcurver", quietly = TRUE)}
fits <- growthcurver::SummarizeGrowthByPlate(gc_input)
head(fits)
```

For gcplyr, `as_gcplyr()` produces the long `Well` / `Time` / `Measurements`
shape (metadata columns included).

```{r gcplyr}
head(as_gcplyr(plate, drop_flagged = TRUE))
```

### Averaging technical replicates

With `tech_rep` designated, `collapse_tech = TRUE` averages the wells of each
technical replicate group at every timepoint. Combined with
`drop_flagged = TRUE`, only clean wells enter the average; otherwise `flagged`
tells you when any contributing well was suspect.

```{r collapse}
pooled <- gr_export(plate, collapse_tech = TRUE, drop_flagged = TRUE)
head(pooled)
```

## One-call Quarto report

`gr_report(plate)` renders the bundled Quarto (`.qmd`) template into a
standalone, self-contained HTML report — plate maps, flagged wells with
reasons, thresholds used, curves, and spatial effects — for keeping alongside
the raw export. It needs the Quarto CLI (bundled with recent RStudio).

```{r report, eval = FALSE}
gr_report(plate, file = "run1_qc.html")
```
